Immutable Source-Mapped Abstract Syntax Tree:\ \\ A Design Pattern for Refactoring Engine APIs

### 2016-07-20 paper.tex
The example in section 4 now uses the same refactoring as sectioin 2.

% TODO: Version with [draft] is for draft mode only
% Also comment/uncomment \today in subtitle when in draft mode
\documentclass[prodmode]{acmlarge}
%\documentclass[draft]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{datetime} % for \today
\usepackage{listings} % for lstlisting environment
\usepackage{placeins} % for \FloatBarrier
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\jb}[1]{\fbox{\texttt{#1}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\kb}[1]{\mbox{\texttt{\small #1}\rule{7.3pt}{0pt}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\lb}[1]{\mbox{\texttt{\small #1}\rule{3.2pt}{0pt}\rule[-1pt]{0pt}{.9em}}}

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  Immutable objects tend
to have simpler interfaces, and they are thread-safe.  Both authors also
emphasize that APIs should be easy to learn (even for programmers who do not
read all of the API documentation) and difficult for clients to misuse. Mutable
ASTs have the advantage of ease-of-learning, but have have all the disadvantages
listed above. They can be valid solutions for simpler refactorings or projects
developed by a small team of people, but become problematic when used for 
nontrivial refactorings or for a public API. For a general, public refactoring
API, immutable trees are less error-prone.

@@

The situation gets a bit worse when changes are made to nested constructs.
Recall the example from Section~\ref{sec:Problem}, which added braces to two
\textit{if} statements, one nested inside the other.  Suppose the outer
\textit{if} statement were modified first: Adding braces to it would change the
position at which the closing brace for the inner \textit{if} statement would
need to be added.  Consider a somewhat simplified example.  Pictorially:

\begin{center}
\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15} \\
$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
\jb{\{}\jb{ }%
\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}%
\jb{ }\jb{\}}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19} \\
\end{center}

\noindent
The body of the inner \textit{if} statement begins at offset~12 in the original
code.  But after adding braces to the outer \textit{if} statement, the body of
the inner \textit{if} statement is shifted to offset~14.

>>>>

% TODO: Version with [draft] is for draft mode only
% Also comment/uncomment \today in subtitle when in draft mode
\documentclass[prodmode]{acmlarge}
%\documentclass[draft,prodmode]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{datetime} % for \today
\usepackage{listings} % for lstlisting environment
\usepackage{placeins} % for \FloatBarrier
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\jb}[1]{\fbox{\texttt{#1}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\kb}[1]{\mbox{\texttt{\small #1}\rule{7.3pt}{0pt}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\lb}[1]{\mbox{\texttt{\small #1}\rule{3.2pt}{0pt}\rule[-1pt]{0pt}{.9em}}}

@@

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  Immutable objects tend
to have simpler interfaces, and they are thread-safe.  Both authors also
emphasize that APIs should be easy to learn (even for programmers who do not
read all of the API documentation) and difficult for clients to misuse. Mutable
ASTs have the advantage of ease-of-learning, but have have all the disadvantages
listed above. They can be valid solutions for simpler refactorings or projects
developed by a small team of people, but become problematic when used for 
nontrivial refactorings or for a public API. For a general, public refactoring
API, immutable trees are much more reliable.

@@

The situation gets a bit worse when changes are made to nested constructs.
Recall the example from Section~\ref{sec:Problem}, which reversed conditionals 
of \textit{if} statements. After the addition of the \textit{!} to the 
condition, the offset of the body is different. Consider a somewhat simplified 
example. Pictorially:

\begin{center}
\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}\jb{ }%
\jb{e}\jb{l}\jb{s}\jb{e}\jb{ }%
\jb{b}\jb{(}\jb{)}\jb{;}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19}\\
$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
\jb{i}\jb{f}\jb{(}\jb{!}\jb{1}\jb{)}\jb{ }%
\jb{b}\jb{(}\jb{)}\jb{;}\jb{ }%
\jb{e}\jb{l}\jb{s}\jb{e}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19}\lb{20}\\
\end{center}

\noindent
The body of the first \textit{if} statement begins at offset~6 in the original
code.  But after adding braces to the outer \textit{if} statement, the body of
the inner \textit{if} statement is shifted to offset~7. This must be taken into
account when the bodies of the \textit{if} and \textit{else} are switched.
